---
title: "EDAmigo Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDAmigo Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
EDAmigo is a package designed to make Exploratory Data Analysis (EDA) simple. It provides novice users with streamlined options for automatically cleaning, processing, and viewing their data. However, it also provides advanced users with the ability to perform EDA functions in an interactive manner.



Begin by loading the EDAmigo package. The package comes with two example datasets: Fires and Finance.
```{r setup}
library(EDAmigo)

# Define the dataframe
df <- fires
```

The package requires all dates and times to be properly classed as 'Date', 'POSIXlt', or 'POSIXct.' Let's check the classes of our dataframe columns.
```{r}
str(df)

```
All of our date columns are type 'Date.' However, there are several special characters, improperly typed columns, and missing values. We can either use the autoClean() function to handle all of these situations at one time, or we can use handleSpecial(), detectTypes(), and handleMissing() individually. Let's use the autoClean() function.

```{r}
# Store output for future use.
# Pass in:  
  # the dataframe, 
  # a user tolerance of 30%, this will prompt the user for direction when a column has <= 30% unique values
  # user_level of 1, indicating that the user prefers to interact with the handleMissing() function

result <- autoClean(df, user_tol=30, user_level = 1)
```




